### 2025-08-05 base-rates.qmd
20250805 - heuristics, biases, and fallacies

In general, people tend to overestimate the likelihood of low base-rate events.
That is, if the base rate of an event or condition—such as schizophrenia—is low (e.g., ~0.5%), people overestimate the likelihood that a person has schizophrenia when given specific information about the person such as their symptoms and history.

@@

In sum, the [marginal probability](#sec-baseRate), including the [prior probability](#sec-baseRate) or [base rate](#sec-baseRate), should be weighed heavily in predictions unless there are sufficient data to indicate otherwise, i.e., to update the posterior probability based on new evidence.
Bayes' theorem specifies how prior beliefs (i.e., [base rate](#sec-baseRate) informations) should be integrated with the [predictive accuracy](#sec-predictiveValidity) of the evidence to make predictions.
It thus provides a powerful tool to [anchor](#sec-heuristicsAnchoringAdjustment) predictions to the [base rate](#sec-baseRate) unless sufficient evidence changes the posterior probability (by updating the evidence and [prior probability](#sec-baseRate)).
In general, you should [anchor](#sec-heuristicsAnchoringAdjustment) your predictions to the [base rate](#sec-baseRate) and [adjust](#sec-heuristicsAnchoringAdjustment) from there.
As noted by @Kahneman2011, if you have doubts about the quality of the evidence for a particular prediction question, keep your predictions close to the [base rate](#sec-baseRate), and modify them only modifying them based on the new information.

>>>>

In general, people tend to overestimate the likelihood of low base-rate events.
That is, if the base rate of an event or condition—such as schizophrenia—is low (e.g., ~0.5%), people overestimate the likelihood that a person has schizophrenia when given specific information about the person such as their symptoms and history.
Moreover, @Fox1998a asked National Basketball Association (NBA) fans to estimate the probability that each of 8 teams in the playoff would win the playoffs.
The median sum of probability judgments for the eight teams was 240% [@Fox1998a; @Kahneman2011] even though, in reality, the true probabilities must sum to 100%.
The finding suggests that people were evaluating each team individually, considering the reasons why that particular team could win, without properly accounting for the total probability.

@@

In sum, the [marginal probability](#sec-baseRate), including the [prior probability](#sec-baseRate) or [base rate](#sec-baseRate), should be weighed heavily in predictions unless there are sufficient data to indicate otherwise, i.e., to update the posterior probability based on new evidence.
People tend to ignore base rates (i.e., [base rate neglect](#sec-fallaciesBaseRate)).
In general, people tend to a) overestimate the likelihood of low base-rate events and b) overweight low base-rate events in their decisions [@Kahneman2011].
Bayes' theorem specifies how prior beliefs (i.e., [base rate](#sec-baseRate) informations) should be integrated with the [predictive accuracy](#sec-predictiveValidity) of the evidence to make predictions.
It thus provides a powerful tool to [anchor](#sec-heuristicsAnchoringAdjustment) predictions to the [base rate](#sec-baseRate) unless sufficient evidence changes the posterior probability (by updating the evidence and [prior probability](#sec-baseRate)).
In general, you should [anchor](#sec-heuristicsAnchoringAdjustment) your predictions to the [base rate](#sec-baseRate) and [adjust](#sec-heuristicsAnchoringAdjustment) from there.
As noted by @Kahneman2011, if you have doubts about the quality of the evidence for a particular prediction question, keep your predictions close to the [base rate](#sec-baseRate), and modify them only modifying them based on the new information.

### 2025-08-05 cognitive-bias.qmd
20250805 - heuristics, biases, and fallacies

In our everyday lives, humans are presented with overwhelming amounts of information.
Because human minds cannot parse every piece of information equally, we tend to take mental shortcuts, called [*heuristics*](#sec-heuristics).
These mental shortcuts can be helpful.
They reduce our mental load and can help us make quick judgments to stay alive or to make complex decisions in the face of uncertainty.
For instance, snap judgments can help us rapidly differentiate between a friend and a foe, for survival.
However, these mental shortcuts can also lead us astray and to make systematic errors in our judgments and predictions.
For instance, [heuristics](#sec-heuristics) can lead us to identify patterns out of randomness.
[*Cognitive biases*](#sec-cognitiveBiases) are systematic errors in thinking.
[*Fallacies*](#sec-fallacies) are forms of flawed reasoning.

Below, we provide examples of [heuristics](#sec-heuristics), [*cognitive biases*](#sec-cognitiveBiases), and [*fallacies*](#sec-fallacies).
These are not exhaustive lists of all the [heuristics](#sec-heuristics), [*cognitive biases*](#sec-cognitiveBiases), and [*fallacies*](#sec-fallacies).
They are some examples of key ones that are relevant fantasy football.

>>>>

In our everyday lives, humans are presented with overwhelming amounts of information.
Because human minds cannot parse every piece of information equally, we tend to take mental shortcuts, called [*heuristics*](#sec-heuristics).
These mental shortcuts can be helpful.
They reduce our mental load and can help us make quick judgments to stay alive or to make complex decisions in the face of uncertainty.
For instance, snap judgments can help us rapidly differentiate between a friend and a foe, for survival.
However, these mental shortcuts can also lead us astray and to make systematic errors in our judgments and predictions.
For instance, [heuristics](#sec-heuristics) can lead us to identify patterns out of randomness.
[*Cognitive biases*](#sec-cognitiveBiases) are systematic errors in thinking.
[*Fallacies*](#sec-fallacies) are forms of flawed reasoning (invoked as an argument).
Fallacies are a characteristic of an argument, whereas cognitive biases are characteristics of the reasoning agent.
Thus, [fallacies](#sec-fallacies) can result from [heuristics](#sec-heuristics) and [cognitive biases](#sec-cognitiveBiases).

Below, we provide examples of [heuristics](#sec-heuristics), [cognitive biases](#sec-cognitiveBiases), and [fallacies](#sec-fallacies).
These are not exhaustive lists of all the [heuristics](#sec-heuristics), [cognitive biases](#sec-cognitiveBiases), and [fallacies](#sec-fallacies).
They are some examples of key ones that are relevant fantasy football.
